---
title: "Statistical Analyses for Faiad et. al (2023) PLoS One"
subtitle: "Temperature affects predation of schistosome-competent snails by a novel invader, the marbled crayfish *Procambarus virginalis*"
author: "Code by: Maurice Goodman, Sara Faiad, & Giulio De Leo"
date: "2023"
output: 
  pdf_document:
    latex_engine: xelatex
monofont: JetBrains Mono
---

```{r setup, include=FALSE}
# Data wrangling and plots
library("tidyverse")
library("GGally")
library("cowplot")

# GLMMs and GAMs
library("glmmTMB")
library("mgcv")
library("car")

# Pairwise comparisons
library("emmeans") 
library("multcomp")
library("multcompView")

## Tidy plotting of tables
library("knitr")

## Save figures
knitr::opts_chunk$set(
  fig.height = 4, fig.width = 6, 
  fig.path = "./figures/",
  dev = "png", dpi = 1000,
  message = FALSE, warning = FALSE
)

## Set ggplot theme options
theme_set(
  theme_classic() + 
    theme(
      axis.text = element_text(size = 10, color = "black"),
      axis.title = element_text(size = 12, face = "bold"),
      strip.text = element_text(size = 12, face = "bold"),
      legend.text = element_text(size = 10), 
      strip.background = element_blank(), 
      strip.placement = "outside"
    )
)

# Colors for weights, and for other figures
wght_col <- viridis::mako(3, begin = 0.2, end = 0.6)

## Function to print model summary tables using knitr nicely
tidy_anova <- function(model) {
  
  model |> 
    car::Anova() |> 
    as.data.frame() |> 
    mutate_if(is.numeric, round, digits = 3) |> 
    mutate(
      ` ` = ifelse(`Pr(>Chisq)` < 0.05, "*", ""),
      `Pr(>Chisq)` = ifelse(`Pr(>Chisq)` < 0.001, "<0.001", `Pr(>Chisq)`), 
    ) |> 
    knitr::kable()
  
}

## Function to print post-hoc tests using knitr nicely
tidy_emmeans <- function(table) {
  
  table |> 
    summary() |> 
    as.data.frame() |> 
    mutate_if(is.numeric, round, digits = 3) |> 
    mutate(
      ` ` = ifelse(p.value < 0.05, "*", ""),
      p.value = ifelse(p.value < 0.001, "<0.001", p.value), 
    ) |> 
    knitr::kable()
  
}

## Custom boxplot function for pairs plot
ggally_boxplot <- function(data, mapping, N = 100, ...){
  ggplot(data, mapping) + geom_boxplot(fill = "white", ..., linewidth = 0.8)
}

## Custom histogram function for pairs plot
ggally_facethist2 <- function (data, mapping, ...) 
{
  mapping <- mapping_color_to_fill(mapping)
  horizontal <- is_horizontal(data, mapping)
  if (!horizontal) {
    mapping <- mapping_swap_x_y(mapping)
  }
  xVal <- mapping_string(mapping$x)
  yVal <- mapping_string(mapping$y)
  mapping$y <- NULL
  p <- ggplot(data = data, mapping)
  p <- p + stat_bin(...)
  if (horizontal) {
    p <- p + facet_grid(paste(yVal, " ~ .", sep = "")) + 
      theme(panel.spacing = unit(0.1, "lines")) + 
      scale_y_continuous(n.breaks = 4)
  }
  else {
    p <- p + facet_grid(paste(". ~", yVal, sep = "")) + 
      theme(panel.spacing = unit(0.1, "lines")) + coord_flip() + 
      scale_y_continuous(n.breaks = 4)
  }
  p <- p + labs(x = xVal, y = yVal)
  p
}

set.seed(10)
```


# Data

Reading in and combining datasets, and filtering out the control data, we expect a total of 960 trials in the data:

  - 600 predation trials
    - Round 1: 6 experiments * 6 (12 hr) trials * 5 temperature * 2 species = 360
    - Round 2: 2 experiments * 6 (12 hr) trials * 5 temperature * 2 species * 2 infection status = 240
    
  - 360 control trials
    - Round 1: 2 experiments * 6 (12 hr) trials * 5 temperature * 2 species = 120
    - Round 2: 2 experiments * 6 (12 hr) trials * 5 temperature * 2 species * 2 infection status = 240

Removing data at time 0 (i.e., the start of the first trial in each experimental run), and excluding pilot trial data ("pilot_ex_un_v_in_1"), we have 960 observations (600 experimental, 360 control), as expected. We remove an additional 70 observations (10 control, 60 experimental):

  - All 10 removed control trials, and 42 removed experimental, are marked "dodgy" for various reasons (usually too many snails or snails were missing)
  - 16 experimental trials where crayfish were observed to be molting
  - 2 experimental trials where consumption was negative, but had not been marked "dodgy" - the notes on these trials, however, indicate that they likely should have been marked "dodgy" but weren't.
  
This leaves 350 control and 540 experimental observations.

```{r read_data}
# Read in data
snail_trials <- read.csv("data/all_trials_snail_digital.csv")
cray_trials <- read.csv("data/all_trials_crayfish_digital.csv")

# Generate an id field to link the two sheets
snail_trials <- snail_trials |> unite(craytrialid, c(week, crayfish_id), remove = FALSE) 
cray_trials <- cray_trials |> unite(craytrialid, c(week, crayfish_id), remove = FALSE)

# Combine datasets
# Remove data with more snails than density, and those that were molting
# There are also 2 values with consumption < 0...
all_data <- cray_trials |>
  dplyr::select(craytrialid, weight, berried) |>
  right_join(snail_trials, by = "craytrialid", multiple = "all") |> 
  filter(time != 0) |>
  dplyr::select(-c(start_time, end_time, collector_name, notes)) |>
  mutate(
    consumption_all = snail_density - snails_remaining - dead_snails,
    p_consumed = consumption_all / (snail_density - dead_snails),
    condition = ifelse(grepl("control", condition), "control", "experimental"),
    snail_species = factor(ifelse(snail_species == "biomph", "Bi. glabrata", "Bu. truncatus")),
    infection_status = factor(infection_status),
    temp = factor(temp, ordered = TRUE), 
    time = factor(time, ordered = TRUE)
  ) |>
  filter(week != "pilot_ex_un_v_in_1" & dodgey_remove != "Y" & 
           molt != "Y" & consumption_all >= 0)

## Experimental data only
exp_data <- all_data |> filter(condition == "experimental")
```

## Generalized pairs plot

The data are more or less balanced across experimental treatments (temperature, species, time, and infection status), except that there are many fewer infected than uninfected snails since only uninfected snails were used in the first round of experiments:

```{r pairs_plot, fig.width = 10, fig.height = 8}
exp_data |> 
  dplyr::select(
    temperature = temp, `infection status` = infection_status, time, 
    `crayfish weight` = weight
  ) |> 
  ggpairs(
    aes(color = exp_data$snail_species),
    lower = list(combo = ggally_facethist2),
    diag = list(continuous = wrap("densityDiag", alpha = 0.6)), 
    upper = list(combo = ggally_boxplot), 
    legend = c(1, 1)
  ) + 
  scale_fill_manual(values = wght_col[2:3]) + 
  scale_color_manual(values = wght_col[2:3]) + 
  theme(
    legend.position = "bottom",
    legend.text = element_text(face = "italic"), 
    axis.text = element_text(size = 9)
  ) + 
  labs(fill = "snail species")
```

## Raw consumption data

Plotting snail consumption against covariates, with centered histograms behind boxplots for discrete covariates, and points for the weight plot. The regression line for the number of snails consumed as a function of crayfish weight is estimated with a GLMM whose random effects structure mirrors the model described below.

```{r boxplot_hist, echo = FALSE, fig.height = 10, fig.width = 10}
plot_box_hist <- function(data, var, width = 3) {
  
  var <- enquo(var)
  
  plot_data <- data |> 
    group_by(snail_species, consumption_all, !!var) |> 
    tally() |> ungroup() |> 
    group_by(snail_species, !!var) |> 
    mutate(
      p = n / sum(n), 
      p = p / max(p)
    )
  
  plot_data |> 
    ggplot(aes(!!var, consumption_all)) + 
    geom_rect(aes(
      xmin = !!var - (width * (p/2)), 
      xmax = !!var + (width * (p/2)),
      ymin = consumption_all - 0.3, 
      ymax = consumption_all + 0.3
    ), fill = "grey80") + 
    facet_wrap(~snail_species, nrow = 1) + 
    geom_boxplot(aes(group = !!var), data = data, 
                 width = width / 5, fill = "white", 
                 alpha = 0.9, outlier.color = NA, 
                 linewidth = 1, color = "black") + 
    scale_x_continuous(breaks = sort(unique(data[[quo_name(var)]]))) + 
    scale_y_continuous(breaks = seq(0, 12, 2)) +
    theme(strip.text = element_text(size = 16, face = "bold.italic"),
          plot.margin = margin(-0.4, 0.5, 0.1, 0, "lines"),
          axis.title.y = element_blank()) + 
    coord_cartesian(clip = "off")
}

## Obtain regression trends and standard error for crayfish weight

weight_glm <- glmmTMB::glmmTMB(
  cbind(consumption_all, snail_density - consumption_all) ~ 
    snail_species * weight + (1 | crayfish_id * week), 
  family = binomial(link = "logit"), 
  data = exp_data
)

seq_range <- function(x, n = 300) seq(x[1], x[2], length.out = n)

weight_curves <- data.frame(
  weight = c(
    seq_range(range(exp_data$weight[exp_data$snail_species == "Bu. truncatus"])), 
    seq_range(range(exp_data$weight[exp_data$snail_species == "Bi. glabrata"]))
  ), 
  snail_species = rep(c("Bu. truncatus", "Bi. glabrata"), each = 300), 
  crayfish_id = NA, 
  week = NA
)

weight_pred <- weight_glm |> 
  predict(weight_curves, allow.new.levels = TRUE, se.fit = TRUE)

weight_curves$consumption_all <- plogis(weight_pred$fit) * 12
weight_curves$se <- weight_pred$se.fit
weight_curves$lower <- plogis(qlogis(weight_curves$consumption_all/12) - weight_curves$se) * 12
weight_curves$upper <- plogis(qlogis(weight_curves$consumption_all/12) + weight_curves$se) * 12

## Construct plots

plots <- list()

plots$temp <- exp_data |> 
  mutate(temp = as.numeric(as.character(temp))) |> 
  plot_box_hist(temp, 4) +
  xlab("temperature (C)")

plots$time <- exp_data |> 
  mutate(time = as.numeric(as.character(time))) |> 
  plot_box_hist(time, 9) +
  xlab("time (hrs)") + 
  theme(strip.text = element_text(color = "white"))

plots$infected <- exp_data |> 
  mutate(infection_status = as.numeric(infection_status)) |> 
  plot_box_hist(infection_status, 0.5) +
  xlab("infection status") + 
  scale_x_continuous(breaks = 1:2, labels = c("uninfected", "infected"), 
                     limits = c(0.5, 2.5)) + 
  theme(strip.text = element_text(color = "white"))

plots$weight <- exp_data |> 
  ggplot(aes(weight, consumption_all)) + 
  geom_point(color = "grey80") + 
  geom_ribbon(aes(ymin = lower, ymax = upper), data = weight_curves, alpha = 0.5, 
              fill = "grey60") +
  geom_line(data = weight_curves, linewidth = 0.8) +
  facet_wrap(~snail_species, nrow = 1) + 
  labs(x = "crayfish weight (g)", y = "No. snails consumed") + 
  scale_y_continuous(breaks = seq(0, 12, 2)) +
  theme(strip.text = element_text(size = 16, color = "white"),
        plot.margin = margin(-0.4, 0.5, 0.1, 0, "lines"),
        axis.title.y = element_blank()) + 
  coord_cartesian(clip = "off")

joined_plot <- cowplot::plot_grid(
  grid::textGrob(expression(bold("No. snails consumed")), rot = 90),
  cowplot::plot_grid(plotlist = plots, ncol = 1, labels = "auto"),
  nrow = 1, rel_widths = c(0.05, 0.95)
) + theme(plot.background = element_rect(fill = "white", color = NA))

joined_plot
```

## Consumption as proportions

Here, plotting not the number consumed, but the proportion consumed (out of the initial snail density, minus the number of snails dead from other causes): 

```{r boxplot_jitter, echo = FALSE, fig.height = 10, fig.width = 10}
plot_box_jitter <- function(data, var, width = 3) {
  
  var <- enquo(var)
  
  data |> 
    ggplot(aes(!!var, p_consumed)) + 
    geom_jitter(color = "grey80", width = 0.3) + 
    geom_boxplot(width = 0.15, outlier.color = NA) +
    facet_wrap(~snail_species, nrow = 1) + 
    scale_y_continuous(breaks = seq(0, 1, 0.25), 
                       labels = c("0", "0.25", "0.5", "0.75", "1")) +
    theme(strip.text = element_text(size = 16, face = "bold.italic"),
          plot.margin = margin(-0.4, 0.5, 0.1, 0, "lines"),
          axis.title.y = element_blank())
}

## Curves and standard errors for weight

weight_glm <- glmmTMB::glmmTMB(
  cbind(consumption_all, snail_density - dead_snails - consumption_all) ~ 
    snail_species * weight + (1 | crayfish_id * week), 
  family = binomial(link = "logit"), 
  data = exp_data
)

seq_range <- function(x, n = 300) seq(x[1], x[2], length.out = n)

weight_curves <- data.frame(
  weight = c(
    seq_range(range(exp_data$weight[exp_data$snail_species == "Bu. truncatus"])), 
    seq_range(range(exp_data$weight[exp_data$snail_species == "Bi. glabrata"]))
  ), 
  snail_species = rep(c("Bu. truncatus", "Bi. glabrata"), each = 300), 
  crayfish_id = NA, 
  week = NA
)

weight_pred <- weight_glm |> 
  predict(weight_curves, allow.new.levels = TRUE, se.fit = TRUE)

weight_curves$p_consumed <- plogis(weight_pred$fit)
weight_curves$se <- weight_pred$se.fit
weight_curves$lower <- plogis(qlogis(weight_curves$p_consumed) - weight_curves$se)
weight_curves$upper <- plogis(qlogis(weight_curves$p_consumed) + weight_curves$se)

## Construct plots 

plots$temp <- exp_data |> 
  plot_box_jitter(temp, 3) +
  xlab("temperature (C)")

plots$time <- exp_data |> 
  plot_box_jitter(time, 5) +
  xlab("time (hrs)") + 
  theme(strip.text = element_text(color = "white"))

plots$infected <- exp_data |> 
  plot_box_jitter(infection_status, 0.2) +
  xlab("infection status") + 
  theme(strip.text = element_text(color = "white"))

plots$weight <- exp_data |> 
  ggplot(aes(weight, p_consumed)) + 
  geom_point(color = "grey80") + 
  geom_ribbon(aes(ymin = lower, ymax = upper), data = weight_curves, alpha = 0.5, 
              fill = "grey60") +
  geom_line(data = weight_curves, linewidth = 0.8) +
  facet_wrap(~snail_species, nrow = 1) + 
  labs(x = "crayfish weight (g)") + 
  scale_y_continuous(breaks = seq(0, 1, 0.25), 
                     labels = c("0", "0.25", "0.5", "0.75", "1")) + 
  theme(strip.text = element_text(size = 16, color = "white"), 
        plot.margin = margin(-0.4, 0.5, 0.1, 0, "lines"),
        axis.title.y = element_blank()) + 
  coord_cartesian(clip = "off")

joined_plot <- cowplot::plot_grid(
  grid::textGrob(expression(bold("proportion of snails consumed")), rot = 90),
  cowplot::plot_grid(plotlist = plots, ncol = 1, labels = "auto"),
  nrow = 1, rel_widths = c(0.05, 0.95)
) + theme(plot.background = element_rect(fill = "white", color = NA))

joined_plot
```

# Generalized linear model 

## Model structure

In each trial, there are 12 snails available for consumption (minus those that died), meaning that the response is (0,12) bounded and the response distribution is binomial. The experiment has three main factors - snail species, infection status, and temperature - with repeated observations of the same crayfish across time steps (12, 24, 36, 48, 60, and 72 hours) and across experimental runs ("week"). We'll treat snail species, infected status, temperature, and time steps as fixed effects (with temperature and time as categorical, rather than continuous predictors), and include crayfish weight as well. All second and third order interactions among the experimental factors are included, as well as an interaction between temperature and weight. 

The model is:

$$
\begin{aligned}
  N_{c,i} &\sim \text{Binomial}(p_i, N_{s, i} - N_{d,i}) \\
  \text{logit}(p_i) &\sim \alpha + \beta X_i + \epsilon_{c} + \epsilon_{w} + \epsilon_{cw} \\
\end{aligned}
$$

where:

  - The response $y_i$ is the number of snails eaten (i.e., *Consumption_all* variable), which is distributed according to a binomial distribution with *n* equal to the initial number of snails ($N_{s,i}$, always 12) minus the number of snails that died $N_{d,i}$.
  - The linear predictors of the log-odds that any given snail is eaten $\text{logit}(p_i)$ are (i) an intercept $\alpha$, (ii) some predictors $X$ (detailed above), and (iii) random intercepts for each crayfish and experimental run / week ($\epsilon_c$ and $\epsilon_w$) and an interaction which allows crayfish intercepts to vary by week ($\epsilon_{cw}$).

## Model fitting

Fitting a generalized linear mixed effects model (GLMM) using `glmmTMB`:

```{r glmm_main}
glmm <- glmmTMB(
  cbind(consumption_all, snail_density - dead_snails - consumption_all) ~
    (snail_species + infection_status + temp + time)^3  + weight * temp +
    (1 | crayfish_id * week),
  family = binomial(link = "logit"),
  data = exp_data
)
```

## Main effects

Wald Chi-square tests for the main effects, showing significant effects of temperature, time, and weight. These are "Type II" tests, which according to the `car` documentation "are calculated according to the principle of marginality, testing each term after all others, except ignoring the term's higher-order relatives" - i.e., the temperature effect is a test of the overall effect of temperature in the model, including any interactions. None of the higher-order interaction terms are significant.

```{r anova_main}
tidy_anova(glmm) # Defined in setup block at start of Rmd
```

## Predicted probability of consumption

One way to plot the model estimates is by obtaining predictions on the response scale, i.e. in terms of the probability of consumption for each snail. Here we're plotting data for uninfected *Bu. truncatus*, but these should look similar for the others. Estimates are plotted for snails at weights of 3, 7, and 11 grams , which are approximately the 10% quantile, mean, and 90% quantile from the observed data, respectively.

```{r Bu_truncatus_predictions}
# Combinations of variable levels to predict on
newdata <- expand.grid(
  snail_species = unique(exp_data$snail_species), 
  infection_status = unique(exp_data$infection_status),
  temp = sort(unique(exp_data$temp)), 
  time = sort(unique(exp_data$time)), 
  crayfish_id = NA, week = NA,
  weight = c(3, 7, 11)
)

# Fitted means and standard errors on log-odds scale
fitted <- predict(glmm, newdata, allow.new.levels = TRUE, se.fit = TRUE)
newdata$fit_logit <- as.numeric(fitted$fit)
newdata$se_logit <- as.numeric(fitted$se.fit)

# Transform to probability scale, compute confidence intervals
newdata$fit_prob <- plogis(newdata$fit_logit)
newdata$lower <- plogis(qnorm(0.025, newdata$fit_logit, newdata$se_logit))
newdata$upper <- plogis(qnorm(0.975, newdata$fit_logit, newdata$se_logit))

# Plot
newdata |> 
  mutate(weight = factor(weight)) |> 
  filter(snail_species == "Bu. truncatus" & infection_status == "uninfected") |> 
  ggplot(aes(temp, color = weight)) + 
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0, 
                position = position_dodge(width = 0.4)) + 
  geom_point(aes(y = fit_prob), position = position_dodge(width = 0.4)) + 
  facet_wrap(~time, labeller = label_both) + 
  labs(x = "temperature (C)", y = "P(snail consumption)") + 
  theme(strip.background = element_blank()) + 
  guides(color = guide_legend(reverse = TRUE), fill = guide_legend(reverse = TRUE)) + 
  scale_color_manual(values = wght_col) + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, 0.25))
```


## Post-hoc tests

### Temperature

Here are the estimated marginal means for probability of consumption at each temperature category. The estimated marginal means will differ somewhat from means computed directly from the data, because when averaging temperature effects over all other experimental factors to compute the estimated marginal means, we do so giving each factor equal weight. This produces estimates for the means within each temperature treatment that we would expect to see from a perfectly balanced experimental design:

```{r temp_means}
kable(emmeans(glmm, data = exp_data, specs = "temp", type = "response"), digits = 3)
```

Computing pairwise comparisons among temperature levels, applying a Tukey correction to the p-values:

```{r temp_pairs, message = FALSE}
# Estimated marginal means for each temp term
temp_emmeans <- emmeans(glmm, data = exp_data, specs = "temp")

## Compute pairwise comparisons
temp_emmeans |> pairs() |> tidy_emmeans()
```

Obtaining the letters corresponding to a compact letters display based on the Tukey pairwise comparisons:

```{r temp_cld}
# Letters and estimated marginal means on response scale
temp_cld <- as.data.frame(cld(temp_emmeans, Letters = letters, type = "response"))
```

To visualize the effect of weight on the temperature dependence of predation, we can fit a second model with `gamm`. This model includes temperature and weight only (not their interaction, as this was not significant):

```{r temp_gamm, message = FALSE}
## Fit model with only temperature smoother and random effects
temp_gamm <- gamm(
  cbind(consumption_all, snail_density - dead_snails - consumption_all) ~ 
    s(temp, k = 4) + weight,
  family = binomial(link = "logit"),
  random = list(crayfish_id = ~ 1 | crayfish_id * week),
  data = mutate(exp_data, temp = as.numeric(as.character(temp)))
)
```

We can obtain predictions across a range of temperatures and at different weights:

```{r temp_curve_fit}
## New data to predict on
newdata <- tibble(
  temp = rep(seq(14, 36, length.out = 500), 3),
  weight = rep(c(3, 7, 11), each = 500)
)

# Obtain fitted means and confidence intervals
newdata$fit <- predict(temp_gamm$gam, newdata)
newdata$se <- predict(temp_gamm$gam, newdata, se.fit = TRUE)$se.fit
newdata$fit_prob <- plogis(newdata$fit)
newdata$lower <- plogis(qnorm(0.025, newdata$fit, newdata$se))
newdata$upper <- plogis(qnorm(0.975, newdata$fit, newdata$se))
newdata$weight <- factor(newdata$weight)
```

Plotting the continuous curves beside discrete estimates, with point estimates, confidence intervals, and Tukey letters superimposed:

```{r temp_diptych, echo = FALSE, fig.width = 8}
temp_peak <- newdata$temp[newdata$fit_prob == max(newdata$fit_prob)]

temp_curves <- newdata |> 
  ggplot(aes(temp, fit_prob)) + 
  geom_vline(xintercept = temp_peak, linetype = "dashed", 
             linewidth = 0.8, lineend = "round") +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = weight), 
              data = newdata, alpha = 0.5) + 
  geom_line(aes(color = weight), data = newdata, linewidth = 0.8) +
  geom_text(aes(color = weight, label = paste(weight, "g")), 
            data = newdata |> filter(temp == max(temp)), 
            nudge_x = 0.25, hjust = 0, fontface = "bold") +
  geom_text(aes(x = temp_peak + 0.75, y = 1, 
                label = paste(round(temp_peak, 1), "C")), 
            hjust = 0, color = "grey20") +
  scale_color_manual(values = wght_col) + 
  scale_fill_manual(values = wght_col) + 
  scale_x_continuous(breaks = seq(15, 35, 5), limits = c(14, 37.2)) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "temperature (C)", color = "weight (g)", fill = "weight (g)") + 
  expand_limits(y = 0) + 
  theme(legend.position = "none", 
        axis.title = element_blank(), 
        axis.text.y = element_blank())

cld_plot <- temp_cld |> 
  ggplot(aes(temp)) + 
  geom_jitter(aes(y = p_consumed), color = "grey80", width = 0.3, 
              data = exp_data) + 
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0, 
                linewidth = 1, lineend = "round") + 
  geom_point(aes(y = prob), size = 3) +
  geom_text(aes(y = asymp.UCL, label = .group), nudge_y = 0.04, 
            fontface = "bold", data = temp_cld |> 
              mutate(.group = trimws(.group)) ) +
  scale_y_continuous(breaks = seq(0, 1, 0.25), 
                     labels = c("0", "0.25", "0.5", "0.75", "1")) + 
  coord_cartesian(clip = "off") + 
  theme(axis.title = element_blank())

temp_joined <- plot_grid(plot_grid(
  grid::textGrob(expression(bold("proportion of snails consumed")), rot = 90), 
  cld_plot, temp_curves, nrow = 1, rel_widths = c(0.05, 1, 1), 
  align = "v", axis = "tb"
), grid::textGrob(expression(bold("temperature (C)"))), 
rel_heights = c(1, 0.05), ncol = 1) + 
  theme(plot.background = element_rect(fill = "white", color = NA))

temp_joined
```


### Time

Computing Tukey letters:

```{r time_emmeans, message = FALSE}
# Estimated marginal means for each temp term
time_emmeans <- emmeans(glmm, data = exp_data, specs = "time")

# Letters and estimated marginal means on response scale
time_cld <- as.data.frame(cld(time_emmeans, Letters = letters, type = "response"))
```

We can sample from the fitted means and standard errors to display the implied linear effect of time, as a way of essentially back-transforming the linear trend given by `contrast(time_emmeans, method = "poly")` to the scale of the observed data:

```{r time_linear}
n <- 10000 # Number of posterior samples

## Sample from estimated means and covariance
time <- as.numeric(as.character(summary(time_emmeans)$time))
time_mu <- summary(time_emmeans)$emmean
time_vcov <- vcov(time_emmeans)
time_samples <- mgcv:::rmvn(n, time_mu, time_vcov)

# Fit regressions, extract coefficients
time_coef_sim <- matrix(NA_real_, nrow = n, ncol = 2)

for (i in 1:n) {
  time_lm <- lm(time_samples[i,] ~ time)
  time_coef_sim[i,] <- coef(time_lm)
}

# Coefficient means and covariance
time_mu_sim <- colMeans(time_coef_sim)
time_vcov_sim <- cov(time_coef_sim)
time_se_sim <- sqrt(diag(time_vcov_sim))

## Fitted means and standard errors for range of temperatures
time_seq <- seq(10, 74, by = 0.1)
t_mm <- cbind(rep(1, length(time_seq)), time_seq)
p_hat <- t_mm %*% time_mu_sim
p_se <- sqrt(diag(t_mm %*% time_vcov_sim %*% t(t_mm)))

## Fitted confidence interval
time_fit <- tibble(
  time = time_seq,
  mean = plogis(p_hat), 
  lower = plogis(qnorm(0.025, p_hat, p_se)),
  upper = plogis(qnorm(0.975, p_hat, p_se))
)

## Equation for regression fit
eqn <- expr(
  log * ~~ bgroup("(", frac(p, 1 - p), ")" ) * phantom(0) == 
    ~~ !!round(time_mu_sim[1], 3) * ~~-~~ !!round(abs(time_mu_sim[2]), 3) %*% t
)
```

Superimposing the estimated means and confidence intervals on the linear tend component and confidence band:

```{r time_plot}
time_cld |> 
  mutate(time = as.numeric(as.character(time)), .group = trimws(.group)) |> 
  ggplot(aes(time)) + 
  geom_ribbon(aes(ymin = lower, ymax = upper), data = time_fit,
              alpha = 0.4, fill = wght_col[3]) + 
  geom_line(aes(y = mean), data = time_fit, color = wght_col[3], linewidth = 0.8) +
  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL), width = 0, 
                linewidth = 1, lineend = "round") + 
  geom_point(aes(y = prob), size = 3) + 
  geom_text(aes(y = asymp.UCL, label = .group), nudge_y = 0.025, fontface = "bold") +
  annotate("text", label = eqn, x = 12, y = 0.05, hjust = 0) +
  scale_x_continuous(breaks = seq(12, 72, 12)) +
  scale_y_continuous(expand = c(0, 0), breaks = seq(0, 0.3, 0.1), 
                     labels = c("0", "0.1", "0.2", "0.3")) + 
  labs(x = "time (hrs)",  y = "probability of snail consumption") + 
  expand_limits(y = 0) + 
  coord_cartesian(clip = "off")
```

We can use these samples to test the linear effect of time:

```{r time_ttest}
round(c(
  slope = time_mu_sim[2], se = time_se_sim[2],
  quantile(time_coef_sim[,2], c(0.025, 0.975)), 
  t = time_mu_sim[2]/time_se_sim[2], 
  p.value = pt(time_mu_sim[2]/time_se_sim[2], df = df.residual(glmm)) * 2
), 4)
```

# Unavaiable snails

Here, we'll use the snail behavior data to examine whether the number of snails unavailable to crayfish was related to the treatments, i.e., whether treatments impacted snail avoidant behavior.

First, we create an unavailable snails variable, dropping NA values for and therefore excluding R1 data in which snail behavior was not evaluated:

```{r}
r2_data <- all_data |> 
  drop_na(c(snails_out, snails_under_shelter)) |> 
  mutate(
    unavailable_snails = snails_out + snails_under_shelter, 
    p_unavailable = unavailable_snails / (snail_density - dead_snails)
  )
```

The mean proportion of unavailable snails in each trial is `r round(mean(r2_data$unavailable_snails/r2_data$snail_density), digits = 3)` (SE = `r round(sd(r2_data$unavailable_snails/r2_data$snail_density)/sqrt(nrow(r2_data)), digits = 3)`). By treatment group (experimental vs. control) and snail species, we have:

```{r}
r2_data |> 
  group_by(snail_species, condition) |> 
  summarize(
    N = n(), 
    `P(unavailable)` = mean(unavailable_snails / snail_density), 
    `SE(P(unavailable))` = sd(unavailable_snails / snail_density) / sqrt(n())
  ) |> 
  kable(digits = 3)
```


## Model

Fitting another binomial model with the number of unavailable snails as a response, and dropping the crayfish random effect as crayfish are absent from control trials:

```{r glmm_unavail}
unavailable_snail_model <- glmmTMB(
  cbind(unavailable_snails, snail_density - dead_snails - unavailable_snails) ~ 
    snail_species * infection_status * condition + (1 | week),
  family = binomial(link = "logit"), 
  data = r2_data
)
```

The null hypothesis tests for the main effects and interactions:

```{r anova_unavail}
tidy_anova(unavailable_snail_model)
```

## Pairwise comparisons

### Main effects

Examining estimated means across the main effects of condition:

```{r unavail_condition}
kable(emmeans(unavailable_snail_model, specs = "condition", type = "response"), digits = 3)
```

and snail species:

```{r unavail_species}
kable(emmeans(unavailable_snail_model, specs = "snail_species", type = "response"), digits = 3)
```


### Interaction

While the third order interaction between snail species, infection status, and condition is not significant, two of the lower order interactions are, so I'll compute pairwise interactions for these treatments together:

```{r unavail_interaction}
unsnail_emmeans <- emmeans(
  unavailable_snail_model, 
  specs = c("snail_species", "infection_status", "condition")
)

unsnail_cld <- as.data.frame(cld(unsnail_emmeans, Letters = letters, type = "response"))
unsnail_cld$.group <- trimws(unsnail_cld$.group)
```

Plotting fitted means and letters:

```{r unavailable_snails_plot}
unsnail_cld |> 
  mutate(condition = as.numeric(as.factor(condition))) |> 
  ggplot(aes(condition)) + 
  geom_jitter(aes(x = as.numeric(as.factor(condition)) - 0.125, y = p_unavailable), 
              data = r2_data |> filter(infection_status == "infected"), 
              alpha = 0.2, width = 0.1, color = colorspace::lighten("#CD8E04", 0.4)) + 
  geom_jitter(aes(x = as.numeric(as.factor(condition)) + 0.125, y = p_unavailable), 
              data = r2_data |> filter(infection_status == "uninfected"), 
              alpha = 0.2, width = 0.1, color = colorspace::lighten("#5E9FC8", 0.4)) + 
  geom_pointrange(aes(y = prob, ymin = asymp.LCL, ymax = asymp.UCL, color = infection_status), 
                  lineend = "round", linewidth = 1, position = position_dodge(0.5), size = 0.7) +
  geom_text(aes(y = asymp.UCL + 0.03, label = .group, color = infection_status), size = 4,
            position = position_dodge(0.5), fontface = "bold", show.legend = FALSE) + 
  facet_wrap(~snail_species, nrow = 1, strip.position = "bottom") + 
  scale_color_manual(name= "Infection Status", labels = c("infected", "uninfected"),
                     values = c("#CD8E04", "#5E9FC8")) + 
  ylab("proportion of snails unavailable") + 
  scale_x_continuous(breaks = 1:2, labels = c("control", "experimental")) +
  theme(
    axis.title.x = element_blank(), 
    strip.text = element_text(face = "italic")
  )
```

# Snail mortality

Summarizing the number of trials with snail mortality and the snail mortality rate (proportion of dead snails) in control and experimental tanks by snail species:

```{r mortality_table}
all_data |> 
  group_by(snail_species, condition) |> 
  summarize(
    N = n(), 
    `trials with mortality` = sum(dead_snails > 0),
    `mortality rate` = mean(dead_snails / snail_density), 
    `SE(mortality rate)` = sd(dead_snails / snail_density) / sqrt(n())
  ) |> 
  kable(digits = 3)
```

## Model

Fitting a similar model to the above, with the number of dead snails as the response and using all 890 observations (350 control and 540 experimental):

```{r glmm_mortality}
dead_snail_model <- glmmTMB(
  cbind(dead_snails, snail_density - dead_snails) ~ 
    snail_species * infection_status * condition + (1 | week),
  family = binomial(link = "logit"), 
  data = all_data
)
```

Null hypothesis tests of the main effects and interactions, suggesting that mortality differed between species, between infected and uninfected snails, between experimental and control tanks, and that the effect of the treatment on snail mortality differed among species:

```{r anova_mortality}
tidy_anova(dead_snail_model)
```

## Pairwise comparisons

### Infection status

There's only two categories here, so the p-value is the p-value of the main effect for infection status reported above, but here are the estimated mortality rates for infected and uninfected snails, averaging over species and condition due to the lack of interactions between infection status and these covariates. While the effect is significant, it's *very* small - a difference of 0.3%:

```{r mortality_status}
kable(emmeans(dead_snail_model, "infection_status", type = "response"), digits = 3)
```


### Condition by species

Examining the snail species by condition interaction, we see that mortality was significantly higher in treatment tanks than in control tanks for both species, but still fairly low overall, with *Bu. truncatus* and *Bi. glabrata* in the experimental having a mortality rate of about 3% and 1%, respectively.

```{r dead_snails_plot}
dsnail_emmeans <- emmeans(dead_snail_model, specs = c("snail_species", "condition"))

dsnail_cld <- as.data.frame(cld(dsnail_emmeans, Letters = letters, type = "response"))
dsnail_cld$.group <- trimws(dsnail_cld$.group)

dsnail_cld |> 
  ggplot(aes(condition)) +
  geom_pointrange(aes(y = prob, ymin = asymp.LCL, ymax = asymp.UCL), lineend = "round", 
                  linewidth = 1, position = position_dodge(0.5), size = 0.7) +
  geom_text(aes(y = asymp.UCL + 0.005, label = .group), position = position_dodge(0.5),
            fontface = "bold", show.legend = FALSE) + 
  facet_wrap(~snail_species, nrow = 1, strip.position = "bottom") + 
  ylab("snail mortality") + 
  theme(
    axis.title.x = element_blank(), 
    strip.text = element_text(face = "italic")
  )
```
